Parallel Adaptive Importance Sampling

### 2016-06-01 PAIS.tex
added description of PAIS-LA and sorted out the names

% modifications to algorithm + spikes
In this problem we modify our PAIS algorithm to use a mixture of Gamma distributions in the proposal distribution instead of a Gaussian mixture. Similarly we use a Gamma proposal in the MH algorithm. This ensures that we only propose positive parameters. Having a heavier right tail in the proposal distribution also means that the posterior is absolutely continuous with respect to the proposal. When this is not the case, the weight function can tend to zero or infinity when $k_i \rightarrow \infty$. Sampling under these conditions can result in very spiky approximations of the tails of the posterior distribution resulting in slow convergence.

@@

% numerics.
\subsubsection{Convergence of PAIS-RW and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}

In the numerics which follow we have used the AMR algorithm to resample with an ensemble size of $M=2500$. The method otherwise remains the same as in previous sections. We perform test runs to find the optimal scaling parameters for both RW proposals and MALA-type proposals. We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.

\begin{figure}[htb]
\centering
\subfigure[Optimal values of the scaling parameter.]{\includegraphics[width=0.45\textwidth]{"figures/C1_ESS"}}
\subfigure[Relative L2 error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
\caption{Convergence of the PAIS-RW and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}. Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.}
\label{fig:C1_ESS}
\end{figure}

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm	& MH-Gam & PAIS-RW & PAIS-LA \\ \hline
	$\beta^*$	& 7.2e-0     & 8.9e-1    & 1.0e-0 \\
	\hline
	\end{tabular}
	\vspace{1mm}
	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithms are optimised using the effective sample size.}
	\label{table:C1_opt_beta}
\end{table}

In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_beta}, however the MALA-type proposal achieves a higher effective sample size with the same ensemble size. Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm. The slightly unstable convergence in the PAIS-RW algorithm is the effect of spikes appearing in the tails of the posterior distribution. This is also what has caused the slightly lower effective sample size.

>>>>

% modifications to algorithm + spikes
In this problem we modify our PAIS algorithm to use a mixture of Gamma distributions in the proposal distribution instead of a Gaussian mixture. The PAIS-LA algorithm uses the Langevin proposal distribution from the standard MALA algorithm with a pertubation taken from a Gamma distribution with mean $X+\frac{1}{2}\beta^2\nabla \log\pi(\cdot)$, and  variance $\beta^2$. Similarly we use a Gamma proposal in the MH algorithm. This ensures that we only propose positive parameters. Having a heavier right tail in the proposal distribution also means that the posterior is absolutely continuous with respect to the proposal. When this is not the case, the weight function can tend to zero or infinity when $k_i \rightarrow \infty$. Sampling under these conditions can result in very spiky approximations of the tails of the posterior distribution resulting in slow convergence.

@@

% numerics.
\subsubsection{Convergence of PAIS-Gamma and PAIS-LA vs MH with Gamma proposals}\label{sec:chem_conv}

In the numerics which follow we have used the AMR algorithm to resample with an ensemble size of $M=2500$. The method otherwise remains the same as in previous sections. We perform test runs to find the optimal scaling parameters for both Gamma proposals and MALA-type proposals. We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.

\begin{figure}[htb]
\centering
\subfigure[Optimal values of the scaling parameter.]{\includegraphics[width=0.45\textwidth]{"figures/C1_ESS"}}
\subfigure[Relative L2 error against number of iterations.]{\includegraphics[width=0.45\textwidth]{"figures/C1_L2"}}
\caption{Convergence of the PAIS-Gamma and PAIS-LA algorithms for the chemical system problem described in Section~\ref{sec:chem}. Implementation described in Sections~\ref{sec:chem_implementation} and \ref{sec:chem_conv}.}
\label{fig:C1_ESS}
\end{figure}

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm	& MH-Gamma & PAIS-Gamma & PAIS-LA \\ \hline
	$\beta^*$	& 2.7e-0     & 9.4e-1    & 1.0e-0 \\
	\hline
	\end{tabular}
	\vspace{1mm}
	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithms are optimised using the effective sample size.}
	\label{table:C1_opt_beta}
\end{table}

In Figure~\ref{fig:C1_ESS} (a) we see that the two PAIS algorithms have similar optimal values of the scaling parameter, also shown in Table~\ref{table:C1_opt_beta}, however the MALA-type proposal achieves a higher effective sample size with the same ensemble size. Convergence after the first 50 million samples is shown in Figure~\ref{fig:C1_ESS} (b) and demonstrates that the PAIS algorithm converges faster than the MH algorithm. The slightly unstable convergence in the PAIS-Gamma algorithm is the effect of spikes appearing in the tails of the posterior distribution. This is also what has caused the slightly lower effective sample size.
